% Optimizing Text Generation with Large Language Models via Speculative Sampling
% Master's thesis -- Petros Sideris
% Department of Electrical and Computer Engineering,
% Democritus University of Thrace (DUTH)
%
% Build:  latexmk -pdf thesis.tex     (latexmk runs biber automatically)

\documentclass[a4paper,12pt]{report}

% --- Language and Computer Modern fonts (LuaLaTeX + fontspec) ----------------
% Built with LuaLaTeX so Greek (e.g. \chapter{Εισαγωγή}) and any UTF-8 render.
% New Computer Modern (NewCM) is the modern, Unicode OpenType incarnation of
% Knuth's Computer Modern -- the classic default (La)TeX look -- with FULL Greek
% (monotonic + polytonic) in every weight, so the chapter titles and babel's
% Greek captions ("Κεφάλαιο", "Βιβλιογραφία") render bold/italic correctly.
% (Plain Latin Modern's OTF carries NO Greek, hence NewCM.)  Loaded by file name
% so the bold/italic faces link unambiguously.  LuaTeX is natively UTF-8.
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{fontspec}
\usepackage{unicode-math}
\setmainfont{NewCM10-Regular.otf}[          % Computer Modern (Unicode), full Greek
  BoldFont       = NewCM10-Bold.otf,
  ItalicFont     = NewCM10-Italic.otf,
  BoldItalicFont = NewCM10-BoldItalic.otf]
\setsansfont{NewCMSans10-Regular.otf}[      % CM sans, full Greek
  BoldFont       = NewCMSans10-Bold.otf,
  ItalicFont     = NewCMSans10-Oblique.otf,
  BoldItalicFont = NewCMSans10-BoldOblique.otf]
\setmonofont{NewCMMono10-Regular.otf}[      % CM typewriter (code listings)
  BoldFont       = NewCMMono10-Bold.otf,
  ItalicFont     = NewCMMono10-Italic.otf,
  BoldItalicFont = NewCMMono10-BoldOblique.otf]
\setmathfont{NewCMMath-Regular.otf}         % Computer Modern math (math Greek incl.)
\usepackage[english,greek]{babel}          % main = Greek; English available
\usepackage{microtype}                     % full micro-typography under LuaTeX

% --- Page layout, headers, and floats ---------------------------------------
\usepackage[a4paper,left=3cm,right=3cm,top=2.5cm,bottom=2.5cm,headheight=15pt]%
           {geometry}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage{booktabs}              % publication-quality tables
\usepackage{enumitem}
\usepackage{siunitx}               % numbers and units
\usepackage{xcolor}
\usepackage{listings}              % source-code listings
\usepackage[autostyle]{csquotes}   % context-sensitive quotation marks

% --- Bibliography (biblatex + biber) ----------------------------------------
\usepackage[backend=biber,style=numeric-comp,sorting=none,giveninits=true]%
           {biblatex}
\addbibresource{thesis.bib}

% --- Hyperlinks and smart cross-references ----------------------------------
% hyperref must be loaded before cleveref.
\usepackage[colorlinks=true,linkcolor=black,citecolor=black,%
            urlcolor={blue!55!black},plainpages=false,pdfpagelabels,%
            hypertexnames=false]{hyperref}
\usepackage{bookmark}              % robust single-pass PDF bookmarks/outlines
\usepackage[capitalise,nameinlink]{cleveref}

\hypersetup{
  pdftitle  = {Optimizing Text Generation with Large Language Models via
               Speculative Sampling},
  pdfauthor = {Petros Sideris},
}

% --- Theorem-like environments ----------------------------------------------
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}{Lemma}[chapter]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{example}{Example}[chapter]
\theoremstyle{remark}
\newtheorem{remark}{Remark}[chapter]

% --- Source-code listing style ----------------------------------------------
\definecolor{codegray}{gray}{0.45}
\definecolor{codekeyword}{rgb}{0.13,0.13,0.55}
\definecolor{codestring}{rgb}{0.00,0.40,0.00}
\lstset{
  basicstyle      = \ttfamily\small,
  keywordstyle    = \color{codekeyword}\bfseries,
  commentstyle    = \color{codegray}\itshape,
  stringstyle     = \color{codestring},
  numberstyle     = \ttfamily\footnotesize\color{codegray},
  numbers         = left,
  numbersep       = 8pt,
  showstringspaces= false,
  breaklines      = true,
  frame           = single,
  rulecolor       = \color{black!30},
  captionpos      = b,
  tabsize         = 2,
}

% --- Running headers --------------------------------------------------------
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\nouppercase{\leftmark}}
\fancyhead[R]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}
\fancypagestyle{plain}{%
  \fancyhf{}%
  \fancyfoot[C]{\thepage}%
  \renewcommand{\headrulewidth}{0pt}%
}

% --- "Month Year" helper for the title page ---------------------------------
\newcommand{\monthyear}{%
  \ifcase\month\or January\or February\or March\or April\or May\or June\or
  July\or August\or September\or October\or November\or December\fi
  \space\number\year}

\begin{document}

% ============================== Title page ==================================
\begin{titlepage}
  \centering
  {\scshape\LARGE Democritus University of Thrace\par}
  \vspace{0.4cm}
  {\scshape\large Department of Electrical and Computer Engineering\par}
  \vfill
  {\large Master's Thesis\par}
  \vspace{1.0cm}
  \rule{\linewidth}{0.4pt}\\[0.7\baselineskip]
  {\huge\bfseries Optimizing Text Generation with Large Language Models
   via Speculative Sampling\par}
  \vspace{0.7\baselineskip}
  \rule{\linewidth}{0.4pt}\par
  \vspace{2.5cm}
  \begin{minipage}[t]{0.45\textwidth}
    \raggedright
    {\itshape Author:}\\[0.3em]
    Petros Sideris
  \end{minipage}\hfill
  \begin{minipage}[t]{0.45\textwidth}
    \raggedleft
    {\itshape Supervisor:}\\[0.3em]
    Prof.\ Pavlos S.\ Efraimidis\\
    {\footnotesize Department of ECE, DUTH}
  \end{minipage}
  \vfill
  {Xanthi, Greece\par}
  {\monthyear\par}
\end{titlepage}

% ============================ Front matter ==================================
\pagenumbering{roman}

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
This thesis investigates the speculative sampling technique,
with the goal of using a small language model (SLM) in place of a large one (LLM) 
when the SLM performs equally well.
The large model is activated only when necessary.
Experiments will be conducted to compare different approaches,
measuring response time and energy consumption
in relation to the baseline large models.
The focus will be on:
(a)~the selection of appropriate models, and
(b)~their impact on both accuracy and efficiency.

\vspace{1.5em}
\noindent\textbf{Keywords:}
speculative sampling, speculative decoding, large language models,
text generation, inference optimization, latency, energy efficiency,
\texttt{llama.cpp}.

\tableofcontents

% =============================== Body =======================================
\clearpage
\pagenumbering{arabic}

\chapter{Εισαγωγή}
This research addresses the inference scheduling problem,
drawing from compiler optimization theory
(e.g., branch prediction, profile-guided optimization, instruction scheduling)
to minimize both latency and energy usage.
These parallels arise from the fact that LLM inference can be viewed as a
computational pipeline,
similar to how compilers treat code as a sequence or graph to optimize
(see \href{https://en.wikipedia.org/wiki/Abstract_syntax_tree}{Abstract Syntax Tree}).
Previous research shows speedups of 2--3\(\times\) from speculative decoding
alone~\cite{LKM23,Che23},
and layering compiler-inspired optimizations could yield even greater
improvements,
especially for edge cases like long sequences or low-acceptance-rate drafts.

\section{Speculative decoding overview}
Speculative decoding functions as follows:
\begin{itemize}
  \item The small LLM generates a series of sequential tokens,
        after which the large LLM evaluates all of these in a single pass.
  \item For each token, if the probability in the large LLM is higher than that
        of the small one, it is accepted directly
        (therefore not affecting the large LLM's statistics).
        If the probability is lower, the likelihood of acceptance is
        proportional to the difference in probabilities.
        This method makes it likely that the token will not be accepted,
        in which case the computation is wasted.
        If the small model performs well,
        we gain a speedup without changing the output,
        but if it performs poorly,
        we waste significant compute resources, slowing down the process.
        \emph{(See \href{https://en.wikipedia.org/wiki/Branch_predictor}{Branch Predictor}.)}
  \item Speculative decoding variants allow tokens to exit the model early if
        the model is confident,
        similar to compiler optimizations like function inlining or dead code
        elimination, which skip unnecessary computations.
  \item Speculative sampling often builds a tree of possible token paths during
        text generation.
        We can apply compiler-style graph transformations
        (e.g., pruning redundant branches via static analysis or common
        subexpression elimination on token probabilities)
        to optimize the tree exploration.
  \item Just as compilers schedule operations for parallel execution,
        speculative sampling could \enquote{schedule} token generation to
        prioritize high-probability paths,
        inspired by compiler techniques for out-of-order execution.
\end{itemize}

\section{Research directions and objectives}
This thesis frames LLM inference through the lens of compiler optimization
and explores the following directions and objectives:
\begin{itemize}
  \item \href{https://en.wikipedia.org/wiki/Scheduling_(computing)}{Scheduling (computing)}.
  \item Compiler-inspired runtime design for speculative text generation with
        multi-model pipelines.
  \item Prediction-and-scheduling runtime with compiler-like heuristics.
  \item Code generation for inference graphs.
  \item SLM--LLM pipeline as an intermediate representation (IR).
  \item Adaptive runtime scheduler (like a JIT) for inference.
  \item Profiling framework that drives scheduling decisions.
  \item Performance/energy trade-offs akin to compiler optimization trade-offs.
  \item Investigate the impact of speculative sampling on LLM performance.
  \item Develop methods to enhance the diversity of generated text.
  \item Propose optimizations to improve generation speed while maintaining
        high-quality output.
\end{itemize}

% ============================= References ===================================
\nocite{*}
\printbibliography[heading=bibintoc]   % localized title -> "Βιβλιογραφία"

\end{document}

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:
